---
title: "Environmental"
execute: 
  warning: false
  error: false
filters: 
  - lightbox
lightbox: auto
bibliography: references.bib
---

```{r}
#| label: data-load
#| include: false
library(here)

source(here::here('R', '00_loadpackages.R'))
source(here::here('R', '02.2_load_wrangle_WQ-and-MET.R'))

load(here('output','data', 'MET.Rdata'))

load(here('output','data', 'pi_wq.Rdata'))
load(here('output','data', 'ss_wq.Rdata'))
load(here('output','data', 'fm_wq.Rdata'))
load(here('output','data', 'pc_wq.Rdata'))
```

The following includes variables that establish the estuarine setting: rainfall, light intensity, water temperature, and salinity. These variables are driven primarily by climatic patterns and largely influence water quality condition.

## Summary

Rainfall totals for the first quarter of 2024 exceeded the quarterly totals from the previous five years with the exception of 2022; however second quarter totals were less (@fig-rainfall). June had the greatest amount of rainfall at 144.5 mm (@fig-rainfall-2), and both January and February had the greatest number of rain days exceeding 10 mm/day (@fig-rainfall-3). 
<!-- Winter storm Finn (unofficially named) brought rainfall and strong winds in [January](https://www.actionnewsjax.com/news/local/strong-storm-cold-front-will-move-across-northeast-florida-southeast-georgia/DNJ3SOGJ35G4FCKJLPJNP67YZY/) to Northeast Florida causing some local schools to release students early and cancel after school activities (@fig-rainfall-1; @fig-rainfall-2). This event is discussed in further detail in the next chapter. -->

The start of Q2 was very dry with little to no rain. Some relief from the drought conditions finally came on May 13-15 when the area received heavy rains and strong winds. But those affects were short-lived. As the drought conditions continued into June the light intensity was high especially mid-day (@fig-lightintensity). 

There were record breaking high temperatures June 7-10. Temperatures reached upwards of 100 degrees with a heat index of 110 degrees and still no rain. Daily average air temperatures were variable during the first part of the year but surprisingly fell within the historical range for the GTMNERR station through June  (@fig-range-atemp). January and March experienced a few temperature drops outside of the average range  (@fig-range-atemp). 

Daily averages in water temperature generally fell within historical ranges at the stations within the first two quarters, however, all stations experienced lower temperature in January and higher highs in Q2 (@fig-range-temp). The changes in daily averages were observed at all four stations, though more pronounced at San Sebastian and Fort Matanzas, the stations closed to inlets (@fig-range-temp-2; @fig-range-temp-3).

Rainfall in January brought salinities lower than their average daily ranges (@fig-range-sal), particularly at the stations farthest from the inlets (@fig-range-sal-1; @fig-range-sal-4). However, the drought conditions in spring made salinity levels rise. Salinities stayed above the average daily range at all stations in June but was most pronounced at Pine Island (@fig-range-sal-1).

## Data

### Rainfall

Data is  presented as quarterly totals, then monthly, and then daily totals as "rain days" which are categorized into Wet (\<10 mm), Very Wet (20-29 mm), and Extremely Wet (\>30 mm) days. 

```{r}
#| label: fig-rainfall
#| echo: false
#| fig.cap: Rainfall at the Pellicer Creek weather station.
#| fig.subcap:
#|    - "Quarterly total precipitation for 2024."
#|    - "Monthly total precipitation for 2024."
#|    - "Daily rainfall totals for days with more than 1mm of day."
#|    - "Hourly rainfall intensity counts by category per month."
#| layout-ncol: 2


# QUARTERLY ----

MET |> 
  SWMPrExtension::seasonal_barplot(param = 'totprcp',
                                   season_grps = list(c(1,2,3), 
                                                      c(4,5,6), 
                                                      c(7,8,9), 
                                                      c(10, 11, 12)),
                                   season_names = c('Jan-Mar', 
                                                    'Apr-Jun', 
                                                    'Jul-Sep', 
                                                    'Oct-Dec'),
                                   hist_rng = c(2003,2024),
                                   hist_avg = TRUE,
                                   converted = F) + 
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 0.6),
        axis.text = element_text(color = "black"))


## MONTHLY ----

MET |> 
  filter(datetimestamp > '2024-01-01 00:15'& datetimestamp < '2024-07-01 00:00') %>%
  # mutate(date = as.Date(datetimestamp),
  #        month = lubridate::month(date, label = T)) %>% 
  mutate(date = lubridate::date(datetimestamp),
         month = lubridate::month(datetimestamp, label = TRUE)) %>% 
  group_by(month) %>%
  summarise(sum = sum(totprcp, na.rm = T)) %>%
  # summarise(mean = mean(sum)) %>%
  ggplot(aes(x = month, y = sum)) +
  geom_col(fill = "#0075AC") +
  geom_hline(yintercept = 69.8, linewidth = 1, linetype = "dashed", color = "grey50") +
  geom_text(aes(label = sum), vjust = 1.5, color = "white") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(color = "black", size = 12)) +
  geom_text(aes(x = "Mar", y = 75), 
            label = "69.8 mm",
            size = 4.5,
            color = "grey50") +
  labs(x = '',
       y = 'Precipitation (mm)',
       caption = "Hortizontal line represents monthly average.")

## DAILY ----

fun_in <- function(x) sum(x, na.rm = TRUE)

MET |> 
  filter(datetimestamp > '2024-01-01 00:15' & datetimestamp < '2024-07-01 00:00') %>%
  aggreswmp(by = "days", FUN = fun_in, params = "totprcp") %>% 
  filter(totprcp >= 1) %>% 
  mutate(date = as.Date(datetimestamp),
         month = lubridate::month(date, label = T),
         prcp_in = totprcp * 0.0393701,
         prcp_in = round(prcp_in, digits = 1),
         wet = case_when(totprcp < 10 ~ "Not Very Wet (<10mm)",
                         totprcp >= 10 & totprcp < 20 ~ "Wet (10-19mm)",
                         totprcp >= 20 & totprcp < 30 ~ "Very Wet (20-29mm)",
                         totprcp > 30 ~ "Extremely Wet (>30mm)"),
         wet = factor(wet, levels = c("Not Very Wet (<10mm)",
                                      "Wet (10-19mm)",
                                      "Very Wet (20-29mm)",
                                      "Extremely Wet (>30mm)"))) %>% 
  ggplot(aes(x = date, y = totprcp)) +
  geom_segment(aes(x = date, xend = date, y = 0, yend = totprcp),
               size = 1) +
  geom_point(aes(color = wet), size =2, bg = 4) +
  khroma::scale_color_okabeito() + # change color codes
  scale_y_continuous(expand = c(0,0), limits = c(0,70)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(color = "black", size = 12)) +
  labs(x = '',
       y = 'Total Daily Precipitation (mm)',
       color = "Amount of Rain")

## INTENSITY ----

# rainfall intensity 
# in mm/hr

MET |>
  filter(datetimestamp > '2024-01-01 00:15'& datetimestamp < '2024-07-01 00:00') %>% 
  aggreswmp(by = "hours", FUN = fun_in, params = "totprcp") %>% 
  rename(intensity = totprcp) %>% 
  mutate(date = lubridate::date(datetimestamp),
         month = lubridate::month(datetimestamp, label = TRUE),
         intensity_cat = case_when(intensity == 0 ~ "No Rain (0 mm/hr)",
                                    intensity > 0 & intensity < 2.5 ~ "Light Rain (<2.5 mm/hr)",
                                    intensity >= 2.5 & intensity < 7.6 ~ "Moderate Rain (2.5-7.6 mm/hr)",
                                    intensity >= 7.6 & intensity < 50 ~ "Heavy Rain (7.6-50 mm/hr)",
                                    intensity > 30 ~ "Intense Rain (>50mm/hr)"),
         intensity_cat = factor(intensity_cat, levels = c("No Rain (0 mm/hr)",
                                                          "Light Rain (<2.5 mm/hr)",
                                                          "Moderate Rain (2.5-7.6 mm/hr)",
                                                          "Heavy Rain (7.6-50 mm/hr)",
                                                          "Intense Rain (>50mm/hr)")
                                )
         ) %>% 
  # group_by(month, intensity_cat) %>%
  # count() %>%
  filter(intensity_cat != "No Rain (0 mm/hr)") %>% 
  ggplot() +
  geom_bar(aes(x = month, fill = intensity_cat)) +
  ggthemes::scale_fill_colorblind() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(color = "black", size = 12)) +
  labs(x = "",
       y = "Amount of Hours",
       fill = "Intensity Category")
```

### Light Intensity

The plot below is interactive. Clicking on elements in the legend (months) will select or deselect months of interest. Additionally, the axis can be adjusted by selecting portions of the plot to isolate.

```{r}
#| label: fig-lightintensity
#| echo: false
#| fig.cap: "Total light intensity by time of day for each month based on photosynthetically active radiation data collected at the Pellicer Creek weather station."

ggplotly(
MET %>% 
  filter(datetimestamp > '2024-01-01 00:15'& datetimestamp < '2024-07-01 00:00') %>% 
  select(datetimestamp, totpar) %>% 
  mutate(month = lubridate::month(datetimestamp, label = T), 
         day = lubridate::day(datetimestamp),
         hour = lubridate::hour(datetimestamp),
         minute = lubridate::minute(datetimestamp),
         time = format(as.POSIXct(datetimestamp), format = "%H:%M")) %>% 
  group_by(month, time) %>% 
  summarize(intensity = sum(totpar, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(time = strptime(time, format = "%H:%M", tz = ""),
         datetime = as.POSIXct(time)) %>% 
  ggplot(aes(x = datetime, y = intensity, group = month)) +
  geom_line(aes(color = month)) +
  scale_x_datetime(date_breaks = "hours", 
                   date_labels = "%H:%M") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.text = element_text(color = "black")) +
  labs(x = "Time of Day (HH:MM)",
       y = "Intensity of Photosynthetically Active Radiation (mMol/m-2)")
)
```

### Temperature

Air and water temperatures are presented as daily averages (blue line) with historic ranges (daily range and average daily range) provided in shades of gray. 

```{r}
#| label: fig-range-atemp
#| echo: false
#| fig.cap: "Historical daily range of air temperature at the SWMP weather station with current year daily average overlaid in blue."

  historical_daily_range(MET, param = 'atemp', target_yr = 2024)
```

```{r}
#| label: fig-range-temp
#| echo: false
#| fig.cap: Historical daily range of water temperature at each SWMP water quality station with current year daily average overlaid in blue.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

historical_daily_range(pi, param = 'temp', target_yr = 2024)

historical_daily_range(ss, param = 'temp', target_yr = 2024)

historical_daily_range(fm, param = 'temp', target_yr = 2024)

historical_daily_range(pc, param = 'temp', target_yr = 2024)
```

### Salinity

```{r}
#| label: fig-range-sal
#| echo: false
#| fig.cap: Historical daily range of salinity at each SWMP water quality station with current year daily average overlaid in blue.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

historical_daily_range(pi, param = 'sal', target_yr = 2024)

historical_daily_range(ss, param = 'sal', target_yr = 2024)

historical_daily_range(fm, param = 'sal', target_yr = 2024)

historical_daily_range(pc, param = 'sal', target_yr = 2024)
```


